---
title: "Concomitant Predictors and Distal Outcomes"
author: "Ali Sanaei"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 6
    fig_height: 4
  rmarkdown::pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 6
    fig_height: 4
    latex_engine: xelatex
vignette: >
  %\VignetteIndexEntry{Concomitant Predictors and Distal Outcomes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
RUN_LIVE <- FALSE

knitr::opts_chunk$set(
  eval     = RUN_LIVE,
  message  = FALSE,
  warning  = FALSE,
  fig.path = "distal-",
  collapse = TRUE,
  comment  = "#>"
)

library(mixLCA)

load_fit <- function(nm) {
  path <- system.file("extdata", nm, package = "mixLCA")
  if (!nzchar(path)) path <- file.path("..", "inst", "extdata", nm)
  if (!file.exists(path)) path <- file.path("inst", "extdata", nm)
  readRDS(path)
}
```

# The partitioned architecture

`mixLCA` enforces a temporal separation between three blocks of variables:

1. **Antecedent concomitant predictors**: variables that come *before*
   class formation and shift the probability of belonging to each
   latent class (e.g., demographics).
2. **Contemporaneous manifest indicators**: variables that *define*
   the classes (the measurement model).
3. **Subsequent distal outcomes**: variables that are *consequences*
   of class membership and should not contaminate it.

Mixing these blocks naively produces double-counted information.
Letting a distal outcome help define classes mechanically inflates
its predictive power; letting a concomitant predictor enter the
measurement model leaks its variance into class meaning. `mixLCA`
keeps the three blocks separate: concomitants enter through a
multinomial logistic regression on class membership, and distals are
estimated *after* the measurement model has converged, under
Bolck-Croon-Hagenaars (BCH) inverse-classification-error weighting.

This vignette walks through both pieces on the embedded
`health_screening` dataset: a synthetic 800-subject screening study
with four continuous biomarkers, an age covariate, and a binary
outcome.

# Setup

```{r data}
data(health_screening)
hs_vars <- c("marker_1", "marker_2", "marker_3", "marker_4")
dim(health_screening)
head(health_screening)
```

```{r data-load, eval = !RUN_LIVE, include = FALSE}
data(health_screening)
hs_vars <- c("marker_1", "marker_2", "marker_3", "marker_4")
```

The dataset has no missing values, so listwise deletion is not needed.

# 1. A naive measurement model (no covariates, no distal)

```{r naive}
hs_naive <- lapply(2:3, function(K) {
  fit_lca(health_screening, continuous = hs_vars, n_classes = K,
          control = lca_control(n_starts = 10),
          verbose = FALSE)
})
names(hs_naive) <- paste0("K", 2:3)
```

```{r naive-load, eval = !RUN_LIVE, include = FALSE}
hs_naive <- load_fit("hs_naive_fits.rds")
```

```{r naive-table, eval = TRUE}
compare_models(hs_naive)
```

The K=2 model has the lowest BIC and a clean interpretation
(low-marker vs. high-marker classes). We work with it below.

```{r naive-profiles, eval = TRUE, fig.cap = "Class-mean profiles, naive K=2."}
plot(hs_naive$K2, type = "profiles")
```

# 2. Adding a concomitant predictor

Age is plausibly antecedent to the marker-class status. We let class
membership depend on age via multinomial logistic regression.

## Character-vector form

The simplest specification passes a character vector of variable
names:

```{r concom-chr}
hs_concom_chr <- fit_lca(
  health_screening, continuous = hs_vars, concomitant = "age",
  n_classes = 2,
  control = lca_control(n_starts = 10),
  verbose = FALSE)
```

```{r concom-chr-load, eval = !RUN_LIVE, include = FALSE}
hs_concom_chr <- load_fit("hs_concom_chr.rds")
```

```{r concom-chr-coefs, eval = TRUE}
hs_concom_chr$concomitant_coefs
```

Each column gives the log-odds (relative to class 1) for the
corresponding non-reference class. The sign and magnitude depend on
the EM's class-label permutation; pair the coefficient with the
class profiles to interpret it.

For inference, pair the coefficients with their standard errors:

```{r concom-chr-se, eval = TRUE}
se <- concomitant_se(hs_concom_chr, health_screening)
data.frame(
  Estimate = round(hs_concom_chr$concomitant_coefs[, 1], 4),
  SE       = round(se[, 1], 4)
)
```

## Formula form

If you want interactions, polynomials, transformations, or factor
dummy-coding, supply a one-sided formula instead. `fit_lca`
constructs the design matrix via `stats::model.matrix`, so anything
that works in `lm()` works here:

```{r concom-fm}
hs_concom_fm <- fit_lca(
  health_screening, continuous = hs_vars,
  concomitant = ~ age + I(age^2),
  n_classes = 2,
  control = lca_control(n_starts = 10),
  verbose = FALSE)
```

```{r concom-fm-load, eval = !RUN_LIVE, include = FALSE}
hs_concom_fm <- load_fit("hs_concom_fm.rds")
```

```{r concom-fm-coefs, eval = TRUE}
hs_concom_fm$concomitant_coefs
```

The quadratic term lets you test whether the age effect saturates.
For these data the quadratic coefficient is effectively zero, which
is consistent with the linear-in-age generative model.

Other useful forms:

```{r concom-fm-examples, eval = FALSE}
# Interaction:
fit_lca(..., concomitant = ~ age * sex)

# Polynomial of degree 3:
fit_lca(..., concomitant = ~ poly(age, 3))

# Spline (requires splines pkg loaded):
fit_lca(..., concomitant = ~ splines::bs(age, df = 4))
```

## NAs in concomitant predictors

`fit_lca()` will refuse to run if any concomitant value is NA. This
is by design: silently dropping rows would desynchronize
`model$posteriors` from your input data and would invalidate any
downstream distal model that expects per-row alignment. Impute or
filter before calling:

```{r na-error, eval = FALSE}
hs_na <- health_screening
hs_na$age[1:3] <- NA
fit_lca(hs_na, continuous = hs_vars, concomitant = "age",
        n_classes = 2)
#> Error: Missing values detected in concomitant predictors.
#> Impute or filter these rows before calling fit_lca() ...
```

# 3. Out-of-sample prediction

`predict()` returns class posteriors for new data using the fitted
parameters. Three output types:

```{r predict-types, eval = TRUE}
new_rows <- health_screening[1:5, ]
prob <- predict(hs_concom_chr, newdata = new_rows)          # default: matrix
cls  <- predict(hs_concom_chr, newdata = new_rows, type = "class")
all  <- predict(hs_concom_chr, newdata = new_rows, type = "all")

list(prob = prob, cls = cls, all = all)
```

If `newdata` has NAs in concomitant columns, the corresponding output
rows are padded with NA so the output length always equals
`nrow(newdata)`:

```{r predict-na, eval = TRUE}
new_with_na <- health_screening[1:5, ]
new_with_na$age[c(2, 4)] <- NA
predict(hs_concom_chr, newdata = new_with_na)
```

Rows 2 and 4 are NA, which makes `cbind(new_with_na, predict(...))`
unambiguous.

# 4. Distal outcome estimation

The screening outcome is a *consequence* of marker class. We do
**not** put it in the measurement model; we estimate it separately
under BCH weighting.

The BCH method (Bolck, Croon, & Hagenaars, 2004; refined with WLS by
Vermunt, 2010, and Bakk, Tekle, & Vermunt, 2013) constructs
class-specific inverse-classification-error weights
from the fitted posteriors, then fits a class-specific regression on
the distal outcome using those weights. Because the measurement model
is frozen before the distal step, no gradient from the distal outcome
can contaminate class meaning.

```{r distal}
hs_distal <- distal(hs_concom_chr, health_screening,
                    formula = outcome ~ age,
                    family  = "binomial")
```

```{r distal-load, eval = !RUN_LIVE, include = FALSE}
hs_distal <- load_fit("hs_distal.rds")
```

```{r distal-print, eval = TRUE}
print(hs_distal)
```

Reading the output: one class has a negative intercept (low baseline
P(outcome = "yes")), the other a less-negative or positive intercept
(higher baseline risk). The `age` slope within each class is small
and positive: class membership absorbs most of the age signal, with a
modest within-class residual age effect.

The **classification error matrix** at the bottom tells you how
confident the modal assignment is. Diagonal dominance close to 1
means the BCH adjustment is mild; values closer to 0.5 mean
classification is uncertain and the BCH weights are pulling more
strongly.

## Other families

`distal()` supports gaussian, binomial, and poisson responses through
the `family` argument:

```{r distal-families, eval = FALSE}
# Continuous distal outcome:
distal(hs_concom_chr, my_data, my_continuous_outcome ~ age,
       family = "gaussian")

# Count distal outcome:
distal(hs_concom_chr, my_data, n_events ~ age, family = "poisson")
```

The IRLS solver inside `distal()` handles negative BCH weights via
eigen-projection plus step-halving (a real divergence risk in earlier
versions) and reports `NA` standard errors when the sandwich
estimator goes negative (rather than the misleading $p \approx 0$ you
get from forcing the variance to zero).

# 5. Penalized covariance (when local dependence is continuous)

For continuous indicators with local dependence, switching from
`"full"` to `"penalized"` covariance with `glassoFast` produces exact
sparsity in the inverse covariance matrix.

```{r penalized}
hs_pen <- fit_lca(
  health_screening, continuous = hs_vars, concomitant = "age",
  n_classes = 2, dependence = "penalized",
  control = lca_control(n_starts = 10),
  verbose = FALSE)
```

```{r penalized-load, eval = !RUN_LIVE, include = FALSE}
hs_pen <- load_fit("hs_penalized.rds")
```

The default `penalty = "auto"` selects a heuristic value from data
scale. To request exactly no shrinkage, set `penalty = 0` explicitly.

```{r penalized-cov, eval = TRUE}
round(hs_pen$continuous_params$covariances[[1]], 2)
```

On this dataset the auto-selected penalty is large enough that every
off-diagonal element is shrunk to exactly zero, leaving a diagonal
covariance. This is the right answer: the markers are conditionally
independent within class by construction, so any nonzero
off-diagonal would be over-fitting.

# 6. Putting it together

A typical analysis arc:

```{r workflow-summary, eval = FALSE}
# 1. Decide K with a naive fit
fits <- lapply(2:5, function(K)
  fit_lca(health_screening, continuous = hs_vars, n_classes = K,
          control = lca_control(n_starts = 10)))
compare_models(fits)

# 2. Add concomitants
fit <- fit_lca(health_screening, continuous = hs_vars,
               concomitant = ~ age,
               n_classes = 2,
               control = lca_control(n_starts = 10))

# 3. Diagnose local dependence
bvr_tests(fit, health_screening)

# 4. Estimate distal under BCH
distal(fit, health_screening, outcome ~ age, family = "binomial")
```

See `vignette("workflow")` for the categorical local-dependence side
of the picture (BVR direct effects and SLD), and `vignette("sld-theory")`
for the math.

# Session info

```{r session, eval = TRUE}
sessionInfo()
```
