---
title: "Concomitant Predictors and Distal Outcomes"
author: "Ali Sanaei"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 6
    fig_height: 4
  rmarkdown::pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 6
    fig_height: 4
    latex_engine: xelatex
vignette: >
  %\VignetteIndexEntry{Concomitant Predictors and Distal Outcomes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
RUN_LIVE <- FALSE

knitr::opts_chunk$set(
  eval     = RUN_LIVE,
  message  = FALSE,
  warning  = FALSE,
  fig.path = "distal-",
  collapse = TRUE,
  comment  = "#>"
)

library(mixLCA)

load_fit <- function(nm) {
  path <- system.file("extdata", nm, package = "mixLCA")
  if (!nzchar(path)) path <- file.path("..", "inst", "extdata", nm)
  if (!file.exists(path)) path <- file.path("inst", "extdata", nm)
  readRDS(path)
}
```

# The partitioned architecture

`mixLCA` enforces a temporal separation between three blocks of variables:

1. **Antecedent concomitant predictors**: variables that come *before*
   class formation and shift the probability of belonging to each
   latent class (e.g., demographics).
2. **Contemporaneous manifest indicators**: variables that *define*
   the classes (the measurement model).
3. **Subsequent distal outcomes**: variables that are *consequences*
   of class membership and should not contaminate it.

Mixing these blocks naively produces double-counted information.
Letting a distal outcome help define classes mechanically inflates
its predictive power; letting a concomitant predictor enter the
measurement model leaks its variance into class meaning. `mixLCA`
keeps the three blocks separate: concomitants enter through a
multinomial logistic regression on class membership, and distals are
estimated *after* the measurement model has converged, under
Bollen-Curran-Heckman (BCH) inverse-classification-error weighting.

This vignette walks through both pieces on the Pima Indians Diabetes
dataset, a small, well-known mixed dataset with continuous health
indicators, an age covariate, and a binary diabetes outcome.

# Setup

```{r data}
data("PimaIndiansDiabetes2", package = "mlbench")
pima_vars <- c("glucose", "pressure", "mass", "pedigree")
pima <- PimaIndiansDiabetes2[stats::complete.cases(
  PimaIndiansDiabetes2[, c(pima_vars, "age", "diabetes")]), ]
dim(pima)
head(pima[, c(pima_vars, "age", "diabetes")])
```

```{r data-load, eval = !RUN_LIVE, include = FALSE}
data("PimaIndiansDiabetes2", package = "mlbench")
pima_vars <- c("glucose", "pressure", "mass", "pedigree")
pima <- PimaIndiansDiabetes2[stats::complete.cases(
  PimaIndiansDiabetes2[, c(pima_vars, "age", "diabetes")]), ]
```

A few hundred complete cases remain after listwise deletion on the
four indicators plus `age` and `diabetes`.

# 1. A naive measurement model (no covariates, no distal)

```{r naive}
pima_naive <- lapply(2:3, function(K) {
  fit_lca(pima, continuous = pima_vars, n_classes = K,
          control = lca_control(n_starts = 10, seed = 110),
          verbose = FALSE)
})
names(pima_naive) <- paste0("K", 2:3)
```

```{r naive-load, eval = !RUN_LIVE, include = FALSE}
pima_naive <- load_fit("pima_naive_fits.rds")
```

```{r naive-table, eval = TRUE}
compare_models(pima_naive)
```

The K=2 model has the lowest BIC and the cleanest interpretation
(a "metabolically healthier" vs. "less healthy" split). We work with it
below.

```{r naive-profiles, eval = TRUE, fig.cap = "Class-mean profiles, naive K=2."}
plot(pima_naive$K2, type = "profiles")
```

# 2. Adding a concomitant predictor

The age of the respondent is plausibly antecedent to her metabolic
class (the data come from women 21+). We let class membership depend
on age via multinomial logistic regression.

## Character-vector form

The simplest specification passes a character vector of variable
names:

```{r concom-chr}
pima_concom_chr <- fit_lca(
  pima, continuous = pima_vars, concomitant = "age",
  n_classes = 2,
  control = lca_control(n_starts = 10, seed = 110),
  verbose = FALSE)
```

```{r concom-chr-load, eval = !RUN_LIVE, include = FALSE}
pima_concom_chr <- load_fit("pima_concom_chr.rds")
```

```{r concom-chr-coefs, eval = TRUE}
pima_concom_chr$concomitant_coefs
```

Each column gives the log-odds (relative to class 1) for the
corresponding non-reference class. Here the coefficient on `age` is
positive, meaning older respondents are more likely to fall in class 2,
consistent with metabolic markers worsening with age.

For inference, pair the coefficients with their standard errors:

```{r concom-chr-se, eval = TRUE}
se <- concomitant_se(pima_concom_chr, pima)
data.frame(
  Estimate = round(pima_concom_chr$concomitant_coefs[, 1], 4),
  SE       = round(se[, 1], 4)
)
```

## Formula form

If you want interactions, polynomials, transformations, or factor
dummy-coding, supply a one-sided formula instead. `fit_lca`
constructs the design matrix via `stats::model.matrix`, so anything
that works in `lm()` works here:

```{r concom-fm}
pima_concom_fm <- fit_lca(
  pima, continuous = pima_vars, concomitant = ~ age + I(age^2),
  n_classes = 2,
  control = lca_control(n_starts = 10, seed = 110),
  verbose = FALSE)
```

```{r concom-fm-load, eval = !RUN_LIVE, include = FALSE}
pima_concom_fm <- load_fit("pima_concom_fm.rds")
```

```{r concom-fm-coefs, eval = TRUE}
pima_concom_fm$concomitant_coefs
```

The quadratic captures whether the age effect saturates. With a single
positive linear term and a small negative quadratic, the model says
class-2 membership rises with age, but the marginal effect declines
above a certain age.

Other useful forms:

```{r concom-fm-examples, eval = FALSE}
# Interaction:
fit_lca(..., concomitant = ~ age * sex)

# Polynomial of degree 3:
fit_lca(..., concomitant = ~ poly(age, 3))

# Spline (requires splines pkg loaded):
fit_lca(..., concomitant = ~ splines::bs(age, df = 4))
```

## NAs in concomitant predictors

`fit_lca()` will refuse to run if any concomitant value is NA. This
is by design: silently dropping rows would desynchronise
`model$posteriors` from your input data and would invalidate any
downstream distal model that expects per-row alignment. Impute or
filter before calling:

```{r na-error, eval = FALSE}
pima_na <- pima
pima_na$age[1:3] <- NA
fit_lca(pima_na, continuous = pima_vars, concomitant = "age",
        n_classes = 2)
#> Error: Missing values detected in concomitant predictors.
#> Impute or filter these rows before calling fit_lca() ...
```

# 3. Out-of-sample prediction

`predict()` returns class posteriors for new data using the fitted
parameters. Three output types:

```{r predict-types, eval = TRUE}
new_rows <- pima[1:5, ]
prob <- predict(pima_concom_chr, newdata = new_rows)          # default: matrix
cls  <- predict(pima_concom_chr, newdata = new_rows, type = "class")
all  <- predict(pima_concom_chr, newdata = new_rows, type = "all")

list(prob = prob, cls = cls, all = all)
```

If `newdata` has NAs in concomitant columns, the corresponding output
rows are padded with NA so the output length always equals
`nrow(newdata)`:

```{r predict-na, eval = TRUE}
new_with_na <- pima[1:5, ]
new_with_na$age[c(2, 4)] <- NA
predict(pima_concom_chr, newdata = new_with_na)
```

Rows 2 and 4 are NA, which makes `cbind(new_with_na, predict(...))`
unambiguous.

# 4. Distal outcome estimation

The diabetes status is a *consequence* of metabolic class. We do
**not** put it in the measurement model; we estimate it separately
under BCH weighting.

The BCH method (Bolck, Croon, Hagenaars 2004; refined by Vermunt
2010) constructs class-specific inverse-classification-error weights
from the fitted posteriors, then fits a class-specific regression on
the distal outcome using those weights. Because the measurement model
is frozen before the distal step, no gradient from the distal outcome
can contaminate class meaning.

```{r distal}
pima_distal <- distal(pima_concom_chr, pima,
                      formula = diabetes ~ age,
                      family  = "binomial")
```

```{r distal-load, eval = !RUN_LIVE, include = FALSE}
pima_distal <- load_fit("pima_distal.rds")
```

```{r distal-print, eval = TRUE}
print(pima_distal)
```

Reading the output: class 1 (the metabolically healthier class) has a
negative diabetes intercept (low baseline risk); class 2 has a
positive intercept (higher baseline risk). The `age` slope within each
class is small. Class membership already absorbs most of the age
signal.

The **classification error matrix** at the bottom tells you how
confident the modal assignment is. Diagonal dominance close to 1 means
the BCH adjustment is mild; values closer to 0.5 mean classification
is uncertain and the BCH weights are pulling more strongly.

## Other families

`distal()` supports gaussian, binomial, and poisson responses through
the `family` argument:

```{r distal-families, eval = FALSE}
# Continuous distal outcome:
distal(pima_concom_chr, pima, outcome ~ age, family = "gaussian")

# Count distal outcome:
distal(pima_concom_chr, pima, n_events ~ age, family = "poisson")
```

The IRLS solver inside `distal()` handles negative BCH weights via
eigen-projection plus step-halving (a real divergence risk before
mixLCA 1.0.1) and reports `NA` standard errors when the sandwich
estimator goes negative (rather than the misleading $p \approx 0$ you
get from forcing the variance to zero).

# 5. Penalized covariance (when local dependence is continuous)

For continuous indicators with local dependence, switching from
`"full"` to `"penalized"` covariance with `glassoFast` produces exact
sparsity in the inverse covariance matrix.

```{r penalized}
pima_pen <- fit_lca(
  pima, continuous = pima_vars, concomitant = "age",
  n_classes = 2, dependence = "penalized",
  control = lca_control(n_starts = 10, seed = 110),
  verbose = FALSE)
```

```{r penalized-load, eval = !RUN_LIVE, include = FALSE}
pima_pen <- load_fit("pima_penalized.rds")
```

The default `penalty = "auto"` selects a heuristic value from data
scale. To request exactly no shrinkage, set `penalty = 0` explicitly.

```{r penalized-cov, eval = TRUE}
round(pima_pen$continuous_params$covariances[[1]], 4)
```

# 6. Putting it together

A typical analysis arc:

```{r workflow-summary, eval = FALSE}
# 1. Decide K with a naive fit
fits <- lapply(2:5, function(K)
  fit_lca(pima, continuous = pima_vars, n_classes = K,
          control = lca_control(n_starts = 10, seed = 110)))
compare_models(fits)

# 2. Add concomitants
fit <- fit_lca(pima, continuous = pima_vars,
               concomitant = ~ age,
               n_classes = 2,
               control = lca_control(n_starts = 10, seed = 110))

# 3. Diagnose local dependence
bvr_tests(fit, pima)

# 4. Estimate distal under BCH
distal(fit, pima, diabetes ~ age, family = "binomial")
```

See `vignette("workflow")` for the categorical local-dependence side
of the picture (BVR direct effects and SLD), and `vignette("sld-theory")`
for the math.

# Session info

```{r session, eval = TRUE}
sessionInfo()
```
