---
title: "Introduction to mixLCA"
author: "Ali Sanaei"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
  rmarkdown::pdf_document:
    toc: true
    latex_engine: xelatex
vignette: >
  %\VignetteIndexEntry{Introduction to mixLCA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

# What `mixLCA` does

`mixLCA` fits finite mixture models for latent class analysis with a
**partitioned architecture**: concomitant predictors enter through a
multinomial logistic regression on class membership; manifest
indicators (continuous, categorical, or mixed) define the measurement
model; distal outcomes are estimated *after* the measurement model
converges, under Bollen-Croon-Hagenaars inverse-classification-error
weighting. The partition prevents distal outcomes from leaking into
class meaning and concomitants from contaminating the measurement
structure.

The package supports two responses to local dependence among manifest
indicators:

- **Direct effects** for categorical items (Hagenaars-Vermunt
  specification search via `auto_bvr()`).
- **Spectral Local Dependence (SLD)** for categorical items: a
  low-rank decomposition of the conditional Burt matrix
  (`auto_sld()`).

For continuous indicators, the covariance structure can be fully
unrestricted, diagonal, or sparsified via graphical lasso
(`dependence = "penalized"`).

# A minimal example

```{r minimal, eval = FALSE}
library(mixLCA)
data("PimaIndiansDiabetes2", package = "mlbench")
pima <- na.omit(PimaIndiansDiabetes2)

fit <- fit_lca(pima,
               continuous  = c("glucose","pressure","mass","pedigree"),
               concomitant = ~ age,
               n_classes   = 2,
               control     = lca_control(n_starts = 10, seed = 110))

summary(fit, data = pima)
plot(fit, type = "profiles")

distal(fit, pima, diabetes ~ age, family = "binomial")
```

# Further reading

The package ships three companion vignettes:

- **`vignette("workflow")`** walks through a categorical LCA from
  scratch on the `poLCA::election` data: naive fit, BVR diagnostics,
  `auto_bvr()` specification search, and SLD comparison. Read this
  first if your data are categorical.

- **`vignette("distal-covariates")`** covers concomitant predictors
  (character vector vs.\ formula), the `predict()` contract with
  `na.action`-style padding, and BCH distal estimation on the Pima
  diabetes data.

- **`vignette("sld-theory")`** explains the mathematics of Spectral
  Local Dependence: the conditional Burt matrix, the rank-$d$
  projector, identification, and the M-step. Read this if you want to
  understand what SLD is doing and when it is the right tool.

# Function index

Core fitting:

- `fit_lca()`, `lca_control()`: entry point and optimiser settings.
- `auto_bvr()`: BVR-guided specification search.
- `auto_sld()`: adaptive Spectral Local Dependence rank selection.
- `enumerate_lca()`: fit a range of $K$ values.

Diagnostics:

- `fit_indices()`: AIC, BIC, aBIC, entropy, ICL.
- `bvr_tests()`, `bvr_categorical()`: bivariate residual diagnostics.
- `class_table()`: modal-class x posterior cross-tabulation.

Inference:

- `predict()` (S3 method): posteriors, modal class, full diagnostic
  frame.
- `distal()`: BCH-weighted regression for distal outcomes.
- `concomitant_se()`, `continuous_se()`: asymptotic standard errors.
- `get_posteriors()`, `get_loadings()`, `coef()`: accessors.

Visualisation:

- `plot()` (S3 method) with `type = c("profiles", "bvr", "distal",
  "uncertainty", "convergence", "categorical", "spectral_scree",
  "spectral_loadings")`.

# Session info

```{r session}
sessionInfo()
```
