---
title: "Spectral Local Dependence: Theory and Worked Example"
author: "Ali Sanaei"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 6
    fig_height: 4
  rmarkdown::pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 6
    fig_height: 4
    latex_engine: xelatex
vignette: >
  %\VignetteIndexEntry{Spectral Local Dependence: Theory and Worked Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
RUN_LIVE <- FALSE

knitr::opts_chunk$set(
  eval     = RUN_LIVE,
  message  = FALSE,
  warning  = FALSE,
  fig.path = "sld-",
  collapse = TRUE,
  comment  = "#>"
)

library(mixLCA)

load_fit <- function(nm) {
  path <- system.file("extdata", nm, package = "mixLCA")
  if (!nzchar(path)) path <- file.path("..", "inst", "extdata", nm)
  if (!file.exists(path)) path <- file.path("inst", "extdata", nm)
  readRDS(path)
}
```

# Why a new method

The local-independence assumption underwrites every classical LCA: given
class $k$, the manifest indicators $Y_1, \ldots, Y_J$ are statistically
independent. In practice this assumption frequently fails. Two
established responses exist.

**Direct effects** (Hagenaars 1988; Vermunt 1999) add a conditional
probability table $P(Y_j \mid Y_{j'}, C = k)$ for one or a few item
pairs $(j, j')$. The remedy is targeted and cheap when local dependence
sits in a small number of identifiable pairs, but it scales as $O(J^2)$
candidate pairs to search and produces dense parameter tables for any
pair admitted.

**Latent trait extensions** (Magidson and Vermunt's mixed mixture
models, Bartolucci's hidden-Markov LCA, factor-mixture models) add a
within-class continuous latent variable. They handle distributed
dependence but break the discrete-class story and introduce numerical
integration.

**Spectral Local Dependence (SLD)** sits between the two. It introduces
a rank-$d_k$ *low-dimensional shift* to the class-conditional log
response surface. The shift is sufficient to capture broad correlated
residuals, but it leaves the class labels discrete and the M-step
closed-form (an eigendecomposition).

This vignette gives the math, the identification argument, the
implementation in `mixLCA`, and a worked example.

# The setup

Let $J$ be the number of categorical items, with item $j$ taking
values in $\{1, \ldots, K_j\}$. Stack the **one-hot encoding** of all
items into a single binary vector
$$
\mathbf{Z}_i \in \{0,1\}^C, \qquad C := \sum_{j=1}^J K_j.
$$
For each respondent $i$, exactly one entry of each item's $K_j$-block
of $\mathbf{Z}_i$ equals 1. Let
$\boldsymbol{\pi}^{(k)} \in [0,1]^C$ be the class-$k$ marginal
response probabilities (the standard LCA parameters), one block per
item summing to 1. Under local independence, the row-wise expectation
of $\mathbf{Z}$ conditional on class $k$ is $\boldsymbol{\pi}^{(k)}$
and the conditional variance has a known multinomial form.

Define the **class-conditional residual**
$$
\mathbf{R}^{(k)}_i := \mathbf{Z}_i - \boldsymbol{\pi}^{(k)}.
$$
Its second moment under the class is the **conditional Burt matrix**
$$
\boldsymbol{\Sigma}^{(k)} :=
  \mathbb{E}\!\left[\mathbf{R}^{(k)}\mathbf{R}^{(k)\top} \mid C = k\right].
$$
Under local independence, the off-block-diagonal entries of
$\boldsymbol{\Sigma}^{(k)}$ are zero: items are uncorrelated given
class. Local dependence shows up as nonzero off-block-diagonal entries.

# The spectral idea

If we modelled $\boldsymbol{\Sigma}^{(k)}$ as a *full* matrix, we would
introduce $\binom{C}{2}$ free off-diagonal parameters per class, far
more than the data support. SLD instead assumes that the off-diagonal
covariance is *low-rank*: there exist $d_k \ll C$ orthonormal vectors
$\mathbf{V}^{(k)} = [\mathbf{v}^{(k)}_1, \ldots, \mathbf{v}^{(k)}_{d_k}]
\in \mathbb{R}^{C \times d_k}$ such that
$$
\boldsymbol{\Sigma}^{(k)} \approx
  \mathbf{V}^{(k)} \boldsymbol{\Lambda}^{(k)} \mathbf{V}^{(k)\top}
$$
where $\boldsymbol{\Lambda}^{(k)} = \mathrm{diag}(\lambda_1^{(k)}, \ldots,
\lambda_{d_k}^{(k)})$ collects the top $d_k$ eigenvalues of
$\boldsymbol{\Sigma}^{(k)}$.

In words: each class has its own dominant directions of residual
covariance, and we model only those. The columns of $\mathbf{V}^{(k)}$
are *spectral loadings*: they tell you which item categories drift
together when conditional independence fails.

# The shifted log-likelihood

We do not parameterize $\boldsymbol{\Sigma}^{(k)}$ directly. Instead we
modify the class-$k$ log-density. Let
$\boldsymbol{\eta}^{(k)}_i := \log \boldsymbol{\pi}^{(k)}$ be the
local-independence log-probability vector. SLD adds a **hollow
projection** to that vector:
$$
\boldsymbol{\eta}^{(k)}_i := \log \boldsymbol{\pi}^{(k)}
  + \mathbf{A}^{\star (k)} \mathbf{R}^{(k)}_i,
\qquad
\mathbf{A}^{\star (k)} := \mathbf{M} \odot \left(
  \mathbf{V}^{(k)} \mathbf{V}^{(k)\top}
\right),
$$
where $\mathbf{M}$ is a $C \times C$ "hollow" mask (1 off-block-diagonal,
0 on-block-diagonal) that prevents the projection from touching the
*within-item* part of the parameters. The within-item part is already
parameterized by $\boldsymbol{\pi}^{(k)}$ and double-counting it would
break identification.

The log-density per item block is then a softmax of the shifted
$\boldsymbol{\eta}^{(k)}_i$ restricted to that item's columns. The
composite log-likelihood is the sum across blocks.

Note: $\mathbf{A}^{\star (k)}$ is symmetric and rank-$d_k$ (before
masking; the mask preserves the rank up to numerical tolerance). The
identification of $\mathbf{V}^{(k)}$ is up to orthogonal rotation
within its column span, the same indeterminacy as in factor analysis or
correspondence analysis.

# Estimation

EM with one twist:

**E-step**. Compute posteriors $\gamma_{ik} \propto \pi_k \cdot
\mathrm{softmax}_J(\boldsymbol{\eta}^{(k)}_i)$, as usual.

**M-step**. For each class $k$:

1. Update marginal response probabilities $\boldsymbol{\pi}^{(k)}$ from
   the weighted posterior counts (standard LCA M-step).
2. Form the weighted conditional residual covariance
   $$
   \widehat{\boldsymbol{\Sigma}}^{(k)} =
     \frac{1}{\sum_i \gamma_{ik}}
     \sum_i \gamma_{ik} \, \mathbf{R}^{(k)}_i \mathbf{R}^{(k)\top}_i,
   $$
   where $\mathbf{R}^{(k)}_i = \mathbf{Z}_i - \boldsymbol{\pi}^{(k)}$.
3. Compute the eigendecomposition of
   $\mathbf{M} \odot \widehat{\boldsymbol{\Sigma}}^{(k)}$, keep the top
   $d_k$ eigenvectors, and form
   $\mathbf{A}^{\star (k)} = \mathbf{M} \odot
   \mathbf{V}^{(k)} \mathbf{V}^{(k)\top}$.
4. Use this $\mathbf{A}^{\star (k)}$ to construct the shifted
   log-density for the next E-step.

The M-step is closed-form (an eigendecomposition), so the per-iteration
cost is $O(K \cdot C^3)$, dominated by eigendecomposition, but with
$C$ rarely exceeding a few dozen this is cheap.

EM monotonicity for this scheme rests on a generalized M-step
argument: the eigendecomposition selects the best rank-$d_k$ projection
in Frobenius norm, which is also the rank-$d_k$ projection that
maximises the Q-function locally. `mixLCA` adds a step-halving
safeguard in [R/04_em_engine.R](../R/04_em_engine.R) that backs off if
a candidate update fails to non-decrease the observed log-likelihood;
in practice the safeguard fires rarely.

# Choosing the ranks $d_1, \ldots, d_K$

Three strategies.

**Fixed rank.** Pass `spectral_rank = 2L` (scalar, same rank for all
classes) or `spectral_rank = c(1L, 2L, 0L)` (length-$K$ vector).

**Pooled basis.** Set `spectral_pool = TRUE` to estimate a single
spectral basis from the class-weighted average of conditional Burt
matrices, applied uniformly to every class. This sacrifices class-
specific dependence structure for a sharper estimate of the dominant
directions, useful when sample size per class is small.

**Adaptive search.** `auto_sld()` runs a greedy forward search: at
each step it identifies the class with the largest unmodelled
eigenvalue, tries adding one rank there, and accepts only if BIC (or
AIC, aBIC, ICL) improves. The search stops when no class can improve
or when a rank cap is reached. The result is class-specific ranks
matched to the data's actual residual structure.

# Identification

A few points worth flagging.

- The marginal probability vector $\boldsymbol{\pi}^{(k)}$ is the
  standard LCA parameter; identification of these probabilities relies
  on the same conditions as classical LCA (sufficient items, sufficient
  class separation).
- The rank-$d_k$ projection $\mathbf{A}^{\star (k)}$ is identified up
  to orthogonal rotation within its column span, exactly as in PCA,
  factor analysis, or multiple correspondence analysis.
- The hollow mask $\mathbf{M}$ is essential. Without it the spectral
  shift would absorb information already encoded in
  $\boldsymbol{\pi}^{(k)}$ (the within-item probabilities), producing
  a non-identified surplus.
- Because the M-step is composite (we maximise the sum of marginal
  item softmaxes, not the full multinomial joint), BIC computed in the
  usual way is *naive*: it under-penalises the composite likelihood.
  For rank selection in publication settings, prefer cross-validation
  over BIC. `mixLCA` flags this in `summary()` output when
  `spectral_rank > 0`.

# Worked example: when SLD wins

The election dataset used in `vignette("workflow")` is dominated by
*pairwise* dependence, so direct effects win there. SLD wins when the
local dependence is broad. Here is a small synthetic example that
illustrates.

We generate $N = 500$ observations from two classes, each with 8
binary items. In class 1, items 1-4 are positively correlated via a
shared latent factor; in class 2 items 5-8 are positively
correlated via a different shared factor. Direct effects would need
to enumerate $\binom{4}{2} \cdot 2 = 12$ pairs to absorb this
structure; SLD captures it with two rank-1 projections.

```{r sim-data, eval = TRUE}
sim_one_class <- function(n, p_base, latent_loading) {
  # latent ~ N(0,1); item j response is 1 with prob expit(logit(p_base[j]) + latent_loading[j] * latent)
  L <- rnorm(n)
  out <- sapply(seq_along(p_base), function(j) {
    eta <- stats::qlogis(p_base[j]) + latent_loading[j] * L
    rbinom(n, 1, plogis(eta))
  })
  as.data.frame(out)
}

withr::with_seed(110, {
  c1 <- sim_one_class(250,
    p_base = c(0.7,0.7,0.7,0.7, 0.3,0.3,0.3,0.3),
    latent_loading = c(2,2,2,2, 0,0,0,0))
  c2 <- sim_one_class(250,
    p_base = c(0.3,0.3,0.3,0.3, 0.7,0.7,0.7,0.7),
    latent_loading = c(0,0,0,0, 2,2,2,2))
  df_sim <- rbind(c1, c2)
})
df_sim[] <- lapply(df_sim, function(x) factor(x, levels = c(0,1), labels = c("no","yes")))
names(df_sim) <- paste0("Q", 1:8)
str(df_sim)
```

```{r fit-sim, eval = TRUE, message = FALSE}
# Naive K=2
fit_naive_sim <- fit_lca(df_sim,
                         categorical = names(df_sim),
                         n_classes = 2,
                         control = lca_control(n_starts = 5, seed = 110),
                         verbose = FALSE)

# SLD K=2 with rank 1 per class
fit_sld_sim <- fit_lca(df_sim,
                       categorical = names(df_sim),
                       n_classes = 2,
                       spectral_rank = c(1L, 1L),
                       control = lca_control(n_starts = 5, seed = 110),
                       verbose = FALSE)

data.frame(
  Model    = c("Naive", "SLD (rank 1 each)"),
  log_lik  = round(c(fit_naive_sim$log_lik, fit_sld_sim$log_lik), 2),
  n_params = c(fit_naive_sim$n_params, fit_sld_sim$n_params),
  BIC      = round(c(fit_indices(fit_naive_sim)$BIC,
                     fit_indices(fit_sld_sim)$BIC), 2)
)
```

The SLD model captures the within-class correlation with one rank
per class and gains BIC despite the extra parameters (naive BIC
$\approx 5350$ vs.\ SLD BIC $\approx 5188$, a savings of about 160
points). Inspecting the loadings shows the rank-1 direction in
class 1 as a bipolar contrast: items Q1-Q4 (the items whose latent
factor was active in class 1) load with one sign, items Q5-Q8 load
with the opposite sign. The single dimension thereby encodes the
covariance pattern induced by the shared latent factor:

```{r sim-loadings, eval = TRUE}
get_loadings(fit_sld_sim)[
  get_loadings(fit_sld_sim)$class == "Class 1" &
    get_loadings(fit_sld_sim)$dimension == 1, ]
```

# When SLD does **not** win

Two failure modes are worth knowing about.

**Pairwise dependence is concentrated in a few item pairs.** Then
direct effects beat SLD because they spend parameters only on the
pairs that need them; SLD has to allocate non-zero coefficients to
every pair that the latent direction projects onto.

**Sample size per class is small** relative to $C$. The conditional
Burt matrix is a $C \times C$ covariance estimate; estimating its
top-$d_k$ eigendecomposition requires enough class-weighted
observations to stabilise the eigenvectors. As a rough rule:
$N_k \gtrsim 10 \cdot C / K$.

In either case `auto_sld()` will tell you by returning
`spectral_rank = 0` for the affected classes: the BIC criterion
will reject the spectral expansion.

# Connection to multiple correspondence analysis

Readers from the categorical-data tradition will notice the kinship
with multiple correspondence analysis (MCA): MCA computes
eigendecompositions of a Burt matrix to embed categorical items in a
Euclidean space. SLD is the *conditional* analogue, applied
class-by-class within a mixture model.

The connection is also computational: MCA's leading dimensions
recover the dominant residual covariance structure of the *pooled*
sample; SLD's per-class eigendecomposition recovers it within each
latent class.

# References

- Bartholomew, D. J., Steele, F., Galbraith, J., & Moustaki, I.
  (2008). *Analysis of Multivariate Social Science Data*. Chapman &
  Hall.
- Hagenaars, J. A. (1988). Latent structure models with direct effects
  between indicators. *Sociological Methods & Research*, 16(3),
  379-405.
- Magidson, J., & Vermunt, J. K. (2001). Latent class factor and
  cluster models, bi-plots, and related graphical displays.
  *Sociological Methodology*, 31, 223-264.
- Vermunt, J. K. (1999). A general nonparametric approach to the
  analysis of ordinal categorical data. *Sociological Methodology*,
  29, 187-223.

The methodological paper for SLD itself is in preparation; this
vignette will be updated with a citation once it is available.

# Session info

```{r session, eval = TRUE}
sessionInfo()
```
