---
title: "A Workflow for Latent Class Analysis: Naive Fit, BVR Adjustments, and Spectral Local Dependence"
author: "Ali Sanaei"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 6
    fig_height: 4
  rmarkdown::pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 6
    fig_height: 4
    latex_engine: xelatex
vignette: >
  %\VignetteIndexEntry{A Workflow for Latent Class Analysis: Naive Fit, BVR Adjustments, and Spectral Local Dependence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
## Toggle this flag to TRUE if you want every chunk to actually run during
## vignette rendering. By default we ship `eval = FALSE` (and load
## pre-computed model fits from inst/extdata/) so building the package
## stays fast, but the pkgdown site and any reader who clones the repo
## can flip this on for a live re-run.
RUN_LIVE <- FALSE

knitr::opts_chunk$set(
  eval     = RUN_LIVE,
  message  = FALSE,
  warning  = FALSE,
  fig.path = "workflow-",
  collapse = TRUE,
  comment  = "#>"
)

library(mixLCA)

## Helpers used in the prose chunks below.
load_fit <- function(nm) {
  ## Two-step lookup: first try the installed package, then the source
  ## tree (useful when rendering vignettes from a not-yet-installed
  ## development checkout).
  path <- system.file("extdata", nm, package = "mixLCA")
  if (!nzchar(path)) path <- file.path("..", "inst", "extdata", nm)
  if (!file.exists(path)) path <- file.path("inst", "extdata", nm)
  readRDS(path)
}
```

# Why this vignette

Latent class models are easy to over-trust. The default specification
assumes the manifest indicators are *locally independent* within each
class. That is, conditional on class membership the items carry no
further information about each other. When that assumption is violated,
the EM algorithm compensates by inventing extra classes whose only
purpose is to absorb pairwise correlation. The resulting solution
inflates $K$, distorts class meaning, and inflates the standard errors
of any downstream distal analysis.

This vignette walks through three responses to that problem on a single
dataset:

1. A **naive fit** that treats local independence as given.
2. A **BVR-guided fit** that adds pairwise direct effects between
   manifest items whose bivariate residuals (Vermunt, 1999) exceed a
   chi-squared threshold.
3. A fit with **Spectral Local Dependence (SLD)**, a low-rank
   decomposition of the conditional Burt matrix that captures
   correlated structure across *many* items at once (see the companion
   vignette `vignette("sld-theory")` for the math).

We use `poLCA::election`, a study of how respondents rated the moral
character, competence, and dishonesty of George W. Bush (`G` items) and
Al Gore (`B` items) in 2000. The items come in conceptually parallel
pairs (`MORALG`/`MORALB`, `KNOWG`/`KNOWB`, and so on), so we have an
*a priori* reason to expect local dependence: ratings of the same
candidate cluster together, and parallel items about the two candidates
may also covary through partisan affect.

# Setup

```{r data}
data("election", package = "poLCA")
cat_items <- c("MORALG", "CARESG", "KNOWG",  "LEADG",  "DISHONG", "INTELG",
               "MORALB", "CARESB", "KNOWB",  "LEADB",  "DISHONB", "INTELB")

# `fit_lca` requires a complete-case rectangle in the indicators (it
# tolerates NAs in the categorical likelihood, but auto_bvr's
# bookkeeping is cleaner without them).
keep <- stats::complete.cases(election[, cat_items])
elec <- election[keep, ]
dim(elec)
```

```{r data-load-precomputed, eval = !RUN_LIVE, include = FALSE}
data("election", package = "poLCA")
cat_items <- c("MORALG", "CARESG", "KNOWG",  "LEADG",  "DISHONG", "INTELG",
               "MORALB", "CARESB", "KNOWB",  "LEADB",  "DISHONB", "INTELB")
keep <- stats::complete.cases(election[, cat_items])
elec <- election[keep, ]
```

Each item is rated on a four-point scale. The respondent count after
listwise deletion is about `r format(sum(keep), big.mark = ",")`.

# 1. The naive fit: ignore local dependence

We fit two-, three-, and four-class models with `dependence = "full"`
on the (nonexistent) continuous side and *implicit* local independence
on the categorical side.

```{r naive-fit}
elec_fits <- lapply(2:4, function(K) {
  fit_lca(elec, categorical = cat_items, n_classes = K,
          control = lca_control(n_starts = 5, seed = 110),
          verbose = FALSE)
})
names(elec_fits) <- paste0("K", 2:4)
```

```{r naive-fit-load, eval = !RUN_LIVE, include = FALSE}
elec_fits <- load_fit("election_naive_fits.rds")
```

```{r naive-compare, eval = TRUE}
compare_models(elec_fits)
```

The standard reading: BIC keeps dropping as we add classes, suggesting
$K \ge 4$. Entropy is in the comfortable range. Most analysts would
declare success here.

That conclusion is fragile.

# 2. Diagnosing local dependence with bivariate residuals

`bvr_categorical()` computes a chi-squared statistic for each pair of
items, contrasting the observed bivariate frequency against what the
model predicts under local independence. Values above 3.84 indicate
significant local dependence ($p < .05$, $\mathrm{df} = 1$).

```{r bvr-table, eval = TRUE}
fit_K3 <- elec_fits$K3
bvr_K3 <- bvr_categorical(fit_K3, elec)
head(bvr_K3, 8)
```

Several pairs sit far above the threshold, and the structure is not
random. The largest residuals fall along the candidate diagonal
(`G`-items pair with `G`-items, `B`-items pair with `B`-items). The
model is forcing parallel ratings of the same candidate to look
conditionally independent when they manifestly are not.

We can also visualise the residual network on the continuous side via
`plot(model, type = "bvr")`; the same diagnostic logic applies to
categorical pairs once you tabulate them.

# 3. BVR-guided specification search

`auto_bvr()` automates the Hagenaars-Vermunt-Magidson specification
search: at each step it identifies the largest BVR pair, adds it as a
direct effect, refits, and stops when BIC ceases to improve (or a cap
is hit).

```{r bvr-auto}
elec_bvr <- auto_bvr(
  data = elec, categorical = cat_items,
  K_range = 3,                 # fix K so the comparison stays clean
  max_direct_effects = 4L,
  bvr_threshold = 3.84,
  seed = 110, verbose = FALSE,
  n_starts = 5)

elec_bvr$auto_path$direct_effects
```

```{r bvr-auto-load, eval = !RUN_LIVE, include = FALSE}
elec_bvr <- load_fit("election_bvr_fit.rds")
```

The search added three direct effects:

```{r bvr-effects-list, eval = TRUE}
elec_bvr$auto_path$direct_effects
```

All three involve the `KNOW` items, which is interpretable: judgements
about how much a candidate "knows" cluster with judgements about
competence (`INTELG`, `LEADB`) on the *same* candidate. The BVR
machinery surfaced what was already visible in the residual table.

```{r bvr-compare, eval = TRUE}
fi_naive <- fit_indices(fit_K3)
fi_bvr   <- fit_indices(elec_bvr)

data.frame(
  Model      = c("Naive K=3", "BVR-adjusted K=3"),
  log_lik    = round(c(fi_naive$log_lik, fi_bvr$log_lik), 2),
  n_params   = c(fi_naive$n_params,     fi_bvr$n_params),
  BIC        = round(c(fi_naive$BIC,    fi_bvr$BIC), 2),
  entropy    = round(c(fi_naive$entropy, fi_bvr$entropy), 3)
)
```

BIC drops sharply (~ 33 points per direct effect on this dataset, far
in excess of the BIC penalty for the extra parameters). The
specification is straightforwardly better, and it does so without
inflating $K$.

# 4. Spectral Local Dependence (SLD)

Direct effects work pair by pair. When the dependence is
*low-dimensional* (for example, an underlying latent direction such
as partisanship that pulls many ratings of the same candidate
together), SLD captures the same structure in a single rank-$d$
projection instead of $d \cdot K$ separate pairwise terms.

`auto_sld()` runs a greedy forward search over class-specific ranks,
adding one rank to the class with the largest unmodelled eigenvalue,
accepting only if BIC improves.

```{r sld-auto}
elec_sld <- auto_sld(
  data = elec, categorical = cat_items,
  n_classes = 3,
  max_rank_per_class = 3L,
  criterion = "BIC",
  seed = 110, verbose = FALSE)
elec_sld$specs$spectral_rank
```

```{r sld-load, eval = !RUN_LIVE, include = FALSE}
elec_sld <- load_fit("election_sld_fit.rds")
```

```{r sld-ranks, eval = TRUE}
elec_sld$specs$spectral_rank
round(fit_indices(elec_sld)$BIC, 2)
```

`auto_sld` settles at ranks `[1, 0, 2]`: class 2 (the
moderate/ambivalent class) gets no spectral structure, while
the two more polarized classes each pick up at least one
low-rank latent direction. The BIC drops from ~34219 (naive K=3) to
~33763, a real gain, though smaller than the BVR-adjusted fit
(~33286).

This comparison is informative on its own:

```{r three-way-compare, eval = TRUE}
fi_naive <- fit_indices(fit_K3)
fi_bvr   <- fit_indices(elec_bvr)
fi_sld   <- fit_indices(elec_sld)

data.frame(
  Model    = c("Naive K=3", "BVR (3 direct effects)", "SLD (ranks 1,0,2)"),
  n_params = c(fi_naive$n_params, fi_bvr$n_params, fi_sld$n_params),
  BIC      = round(c(fi_naive$BIC, fi_bvr$BIC, fi_sld$BIC), 2),
  entropy  = round(c(fi_naive$entropy, fi_bvr$entropy, fi_sld$entropy), 3)
)
```

On this dataset, **pairwise direct effects win**. The election items
divide cleanly into G-pairs and B-pairs, so the local dependence
shows up as a small number of large pairwise residuals, exactly
what direct effects were designed to absorb. SLD captures the same
structure but spends parameters on every item-pair coupling that
the latent direction implies, even pairs that the data does not
flag as dependent.

**SLD shines elsewhere**: when residual covariance is *broadly*
distributed across many items (say, a partisanship dimension that
shifts every rating slightly), the rank-$d$ projection beats $\binom{J}{2}$
candidate direct-effect terms. See `vignette("sld-theory")` for the
math and a worked example.

You can also inspect the SLD loadings to read off which items
contribute to each latent direction:

```{r sld-loadings, eval = TRUE, fig.cap = "SLD loadings: rank 1 in class 3.", out.width = "85%"}
plot(elec_sld, type = "spectral_loadings", dimension = 1, class = 3)
```

# 5. Substantive comparison: who are the classes?

```{r class-props, eval = TRUE}
props_naive <- round(colMeans(get_posteriors(fit_K3)), 3)
props_bvr   <- round(colMeans(get_posteriors(elec_bvr)), 3)
data.frame(class = 1:3, naive = props_naive, bvr_adjusted = props_bvr)
```

The class proportions move only slightly. The substantive class meaning
changes more visibly in the response profiles:

```{r class-profile-naive, eval = TRUE, fig.cap = "Naive K=3: P(rating = 'Extremely well' | class) for the six G-items.", out.width = "85%"}
# Build a small profile plot manually from the categorical params.
# This keeps the vignette dependency-light and shows readers exactly
# how to extract any quantity from a fitted mixLCA object.
extract_extremely_well <- function(fit, items) {
  K <- fit$n_classes
  cp <- fit$categorical_params
  out <- matrix(NA_real_, K, length(items), dimnames = list(paste("Class", 1:K), items))
  for (k in 1:K) {
    for (it in items) {
      probs <- cp[[k]][[it]]
      hit   <- grep("Extremely well", names(probs), ignore.case = TRUE, fixed = FALSE)
      if (length(hit)) out[k, it] <- probs[[hit[1]]]
    }
  }
  out
}

g_items <- c("MORALG","CARESG","KNOWG","LEADG","DISHONG","INTELG")
mat_naive <- extract_extremely_well(fit_K3, g_items)

matplot(t(mat_naive), type = "b", pch = 19, lty = 1,
        xaxt = "n", xlab = "", ylab = "P(rating = 'Extremely well')",
        main = "Naive K=3, items about Bush")
axis(1, at = seq_along(g_items), labels = g_items, las = 2, cex.axis = 0.8)
legend("topright", legend = rownames(mat_naive), pch = 19, col = 1:nrow(mat_naive), bty = "n")
```

The three classes separate as you would expect: one strongly positive
about Bush across all items, one negative, one ambivalent. The BVR-
adjusted model recovers the same broad story but with sharper edges
because direct effects relieve the latent classes from explaining
the within-candidate covariance:

```{r class-profile-bvr, eval = TRUE, fig.cap = "BVR-adjusted K=3: same items.", out.width = "85%"}
mat_bvr <- extract_extremely_well(elec_bvr, g_items)
matplot(t(mat_bvr), type = "b", pch = 19, lty = 1,
        xaxt = "n", xlab = "", ylab = "P(rating = 'Extremely well')",
        main = "BVR-adjusted K=3, items about Bush")
axis(1, at = seq_along(g_items), labels = g_items, las = 2, cex.axis = 0.8)
legend("topright", legend = rownames(mat_bvr), pch = 19, col = 1:nrow(mat_bvr), bty = "n")
```

# 6. Takeaways

- **Run the diagnostic.** A satisfying BIC trajectory across $K$ is not
  evidence that your specification is correct. Always inspect
  `bvr_categorical()` (for categorical indicators) or `bvr_tests()`
  (for continuous) before naming the classes.
- **Pairwise vs. low-rank.** Direct effects handle a few misbehaving
  pairs efficiently; SLD handles broad residual structure efficiently.
  The two are complementary; `fit_lca()` lets you supply both.
- **Don't add classes to absorb dependence.** The naive K=4 fit looked
  like it was winning on BIC. Once we model the local dependence
  properly, K=3 is fine, and class meaning is preserved across
  refits.

See also:

- `vignette("distal-covariates")` for distal outcome estimation under
  BCH weighting, and for the formula interface to concomitant
  predictors.
- `vignette("sld-theory")` for the math and a worked example where
  SLD outperforms BVR.

# Session info

```{r session, eval = TRUE}
sessionInfo()
```
