---
title: "A Workflow for Latent Class Analysis: Naive Fit, BVR Adjustments, and Spectral Local Dependence"
author: "Ali Sanaei"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 6
    fig_height: 4
  rmarkdown::pdf_document:
    toc: true
    toc_depth: 3
    fig_width: 6
    fig_height: 4
    latex_engine: xelatex
vignette: >
  %\VignetteIndexEntry{A Workflow for Latent Class Analysis: Naive Fit, BVR Adjustments, and Spectral Local Dependence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
## Toggle this flag to TRUE if you want every chunk to actually run during
## vignette rendering. By default we ship `eval = FALSE` (and load
## pre-computed model fits from inst/extdata/) so building the package
## stays fast, but the pkgdown site and any reader who clones the repo
## can flip this on for a live re-run.
RUN_LIVE <- FALSE

knitr::opts_chunk$set(
  eval     = RUN_LIVE,
  message  = FALSE,
  warning  = FALSE,
  fig.path = "workflow-",
  collapse = TRUE,
  comment  = "#>"
)

library(mixLCA)

## Helpers used in the prose chunks below.
load_fit <- function(nm) {
  ## Two-step lookup: first try the installed package, then the source
  ## tree (useful when rendering vignettes from a not-yet-installed
  ## development checkout).
  path <- system.file("extdata", nm, package = "mixLCA")
  if (!nzchar(path)) path <- file.path("..", "inst", "extdata", nm)
  if (!file.exists(path)) path <- file.path("inst", "extdata", nm)
  readRDS(path)
}
```

# Why this vignette

Latent class models are easy to over-trust. The default specification
assumes the manifest indicators are *locally independent* within each
class. That is, conditional on class membership the items carry no
further information about each other. When that assumption is violated,
the EM algorithm compensates by inventing extra classes whose only
purpose is to absorb pairwise correlation. The resulting solution
inflates $K$, distorts class meaning, and inflates the standard errors
of any downstream distal analysis.

This vignette walks through three responses to that problem on a single
dataset:

1. A **naive fit** that treats local independence as given.
2. A **BVR-guided fit** that adds pairwise direct effects between
   manifest items whose bivariate residuals (Vermunt, 1999) exceed a
   chi-squared threshold.
3. A fit with **Spectral Local Dependence (SLD)**, a low-rank
   decomposition of the conditional Burt matrix that captures
   correlated structure across *many* items at once (see the companion
   vignette `vignette("sld-theory")` for the math).

We use the package's embedded dataset `voter_perceptions`, a 2000-
respondent synthetic survey rating two hypothetical candidates ("A"
and "B") on six attributes each (moral character, competence,
leadership, honesty, intelligence, empathy). Items come in conceptually
parallel pairs (`moral_A`/`moral_B`, and so on), so we have an *a
priori* reason to expect local dependence: ratings of the same
candidate cluster together via a shared partisan affect.

# Setup

```{r data}
data(voter_perceptions)
cat_items <- names(voter_perceptions)
length(cat_items)   # 12 items
nrow(voter_perceptions)
```

```{r data-load-precomputed, eval = !RUN_LIVE, include = FALSE}
data(voter_perceptions)
cat_items <- names(voter_perceptions)
```

Each item is rated on a four-level ordinal scale (`poor`, `fair`,
`good`, `excellent`).

# 1. The naive fit: ignore local dependence

We fit two-, three-, and four-class models with local independence on
the categorical side.

```{r naive-fit}
voter_fits <- lapply(2:4, function(K) {
  fit_lca(voter_perceptions, categorical = cat_items, n_classes = K,
          control = lca_control(n_starts = 5, seed = 110),
          verbose = FALSE)
})
names(voter_fits) <- paste0("K", 2:4)
```

```{r naive-fit-load, eval = !RUN_LIVE, include = FALSE}
voter_fits <- load_fit("voter_naive_fits.rds")
```

```{r naive-compare, eval = TRUE}
compare_models(voter_fits)
```

The standard reading: BIC drops as we add classes, with $K = 4$
narrowly best. Entropy is in the comfortable range. Most analysts
would declare success around $K = 3$ or $K = 4$.

That conclusion is fragile.

# 2. Diagnosing local dependence with bivariate residuals

`bvr_categorical()` computes a chi-squared statistic for each pair of
items, contrasting the observed bivariate frequency against what the
model predicts under local independence. Values above 3.84 indicate
significant local dependence ($p < .05$, $\mathrm{df} = 1$).

```{r bvr-table, eval = TRUE}
fit_K3 <- voter_fits$K3
bvr_K3 <- bvr_categorical(fit_K3, voter_perceptions)
head(bvr_K3, 8)
```

Several pairs sit far above the threshold, and the structure is not
random. Most of the largest residuals fall along the candidate
diagonal (`_A` items pair with other `_A` items, `_B` items pair with
other `_B` items). The model is forcing parallel ratings of the same
candidate to look conditionally independent when they manifestly are
not.

# 3. BVR-guided specification search

`auto_bvr()` automates the Hagenaars-Vermunt-Magidson specification
search: at each step it identifies the largest BVR pair, adds it as a
direct effect, refits, and stops when BIC ceases to improve (or a cap
is hit).

```{r bvr-auto}
voter_bvr <- auto_bvr(
  data = voter_perceptions, categorical = cat_items,
  K_range = 3,                 # fix K so the comparison stays clean
  max_direct_effects = 4L,
  bvr_threshold = 3.84,
  seed = 110, verbose = FALSE,
  n_starts = 5)
```

```{r bvr-auto-load, eval = !RUN_LIVE, include = FALSE}
voter_bvr <- load_fit("voter_bvr_fit.rds")
```

```{r bvr-effects-list, eval = TRUE}
voter_bvr$auto_path$direct_effects
```

The search added **zero direct effects**. The top candidate
(`honest_B → empath_B`, BVR ~ 39) was tested first; adding it raised
BIC rather than lowering it, so `auto_bvr` rejected it and stopped.

That is informative on its own. The bivariate residuals are
substantively large, but the *cost* of fitting a full conditional
probability table for each pair exceeds the *gain* in fit. The
underlying dependence is broad rather than concentrated in a few
isolated pairs, so pairwise direct effects are not a parameter-
efficient response.

```{r bvr-compare, eval = TRUE}
fi_naive <- fit_indices(fit_K3)
fi_bvr   <- fit_indices(voter_bvr)

data.frame(
  Model      = c("Naive K=3", "auto_bvr K=3"),
  log_lik    = round(c(fi_naive$log_lik, fi_bvr$log_lik), 2),
  n_params   = c(fi_naive$n_params,     fi_bvr$n_params),
  BIC        = round(c(fi_naive$BIC,    fi_bvr$BIC), 2),
  entropy    = round(c(fi_naive$entropy, fi_bvr$entropy), 3)
)
```

Identical numbers because `auto_bvr` returned the naive K=3 model
unchanged. On this dataset, pairwise remediation does not help.

# 4. Spectral Local Dependence (SLD)

When pairwise dependence is broad, the right tool is a *low-rank*
shift to the class-conditional log response surface rather than a
sequence of pairwise terms. `auto_sld()` runs a greedy forward search
over class-specific ranks, adding one rank to the class with the
largest unmodelled eigenvalue, accepting only if BIC improves.

```{r sld-auto}
voter_sld <- auto_sld(
  data = voter_perceptions, categorical = cat_items,
  n_classes = 3,
  max_rank_per_class = 3L,
  criterion = "BIC",
  seed = 110, verbose = FALSE)
```

```{r sld-load, eval = !RUN_LIVE, include = FALSE}
voter_sld <- load_fit("voter_sld_fit.rds")
```

```{r sld-ranks, eval = TRUE}
voter_sld$specs$spectral_rank
round(fit_indices(voter_sld)$BIC, 2)
```

`auto_sld` settles at ranks `[1, 0, 0]`: a single low-rank direction
captures the dominant residual structure in the largest class, while
the smaller classes need no spectral correction. The BIC drops from
~61256 (naive K=3) to ~60997, a savings of ~260 points.

# 5. Three-way comparison

```{r three-way-compare, eval = TRUE}
fi_naive4 <- fit_indices(voter_fits$K4)
fi_sld    <- fit_indices(voter_sld)

data.frame(
  Model    = c("Naive K=3", "Naive K=4", "auto_bvr K=3",
               "SLD K=3 (ranks 1,0,0)"),
  n_params = c(fi_naive$n_params, fi_naive4$n_params,
               fi_bvr$n_params,   fi_sld$n_params),
  BIC      = round(c(fi_naive$BIC, fi_naive4$BIC,
                     fi_bvr$BIC, fi_sld$BIC), 2),
  entropy  = round(c(fi_naive$entropy, fi_naive4$entropy,
                     fi_bvr$entropy, fi_sld$entropy), 3)
)
```

Reading the table: the naive K=4 model and the SLD K=3 model are
roughly tied on BIC. The naive K=4 fit absorbs the residual
covariance by adding an extra class whose only purpose is to soak up
shared variance; the SLD K=3 fit captures the same structure in a
single low-rank direction without inflating the class count. The SLD
solution is more interpretable because the three classes retain their
substantive meaning (favours A, favours B, moderate), and the
residual covariance is explicitly modelled rather than hidden inside
an extra class.

You can inspect the loadings to read off which items contribute to
each latent direction:

```{r sld-loadings, eval = TRUE, fig.cap = "SLD loadings: rank 1 in the largest class.", out.width = "85%"}
plot(voter_sld, type = "spectral_loadings", dimension = 1, class = 1)
```

# 6. Substantive comparison: who are the classes?

```{r class-props, eval = TRUE}
props_naive <- round(colMeans(get_posteriors(fit_K3)), 3)
props_sld   <- round(colMeans(get_posteriors(voter_sld)), 3)
data.frame(class = 1:3, naive = props_naive, sld = props_sld)
```

The class proportions move only slightly between the naive K=3 fit
and the SLD-augmented fit. The substantive class meaning is preserved
across the two specifications:

```{r class-profile-naive, eval = TRUE, fig.cap = "Naive K=3: P(rating = 'excellent') for the six A-items.", out.width = "85%"}
extract_excellent <- function(fit, items) {
  K <- fit$n_classes
  cp <- fit$categorical_params
  out <- matrix(NA_real_, K, length(items),
                dimnames = list(paste("Class", 1:K), items))
  for (k in 1:K) {
    for (it in items) {
      probs <- cp[[k]][[it]]
      hit   <- which(names(probs) == "excellent")
      if (length(hit)) out[k, it] <- probs[[hit[1]]]
    }
  }
  out
}

a_items <- grep("_A$", cat_items, value = TRUE)
mat_naive <- extract_excellent(fit_K3, a_items)

matplot(t(mat_naive), type = "b", pch = 19, lty = 1,
        xaxt = "n", xlab = "", ylab = "P(rating = 'excellent')",
        main = "Naive K=3, items about candidate A")
axis(1, at = seq_along(a_items), labels = a_items, las = 2, cex.axis = 0.8)
legend("topright", legend = rownames(mat_naive), pch = 19,
       col = 1:nrow(mat_naive), bty = "n")
```

The three classes separate as expected from the generative model: one
strongly positive about candidate A, one negative (these are the
class that favours B), and one moderate.

# 7. Takeaways

- **Run the diagnostic.** A satisfying BIC trajectory across $K$ is not
  evidence that your specification is correct. Always inspect
  `bvr_categorical()` (for categorical indicators) or `bvr_tests()`
  (for continuous) before naming the classes.
- **Pairwise vs. low-rank.** Direct effects handle a few large isolated
  residual pairs efficiently; SLD handles broad residual covariance
  efficiently. The two are complementary. `fit_lca()` lets you supply
  both via `cat_direct_effects` and `spectral_rank`.
- **Don't add classes to absorb dependence.** The naive K=4 fit looked
  like it was winning on BIC. Once we model the local dependence
  properly via SLD, K=3 is fine, and class meaning is preserved across
  refits.

See also:

- `vignette("distal-covariates")` for distal outcome estimation under
  BCH weighting, and for the formula interface to concomitant
  predictors.
- `vignette("sld-theory")` for the math and a worked example.

# Session info

```{r session, eval = TRUE}
sessionInfo()
```
